---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# lime
[![Travis-CI Build Status](https://travis-ci.org/thomasp85/lime.svg?branch=master)](https://travis-ci.org/thomasp85/lime)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/thomasp85/lime?branch=master&svg=true)](https://ci.appveyor.com/project/thomasp85/lime)
[![CRAN_Release_Badge](http://www.r-pkg.org/badges/version-ago/lime)](https://CRAN.R-project.org/package=lime)
[![CRAN_Download_Badge](http://cranlogs.r-pkg.org/badges/lime)](https://CRAN.R-project.org/package=lime)

*This is an R port of the Python lime package (https://github.com/marcotcr/lime)
developed by the authors of the lime (Local Interpretable Model-agnostic 
Explanations) approach for black-box model explanations. All credits goes to the
original developers.*

The purpose of `lime` is to explain the predictions of black box classifiers. 
What this means is that for any given prediction and any given classifier it is
able to determine a small set of features in the original data that has driven
the outcome of the prediction. To learn more about the methodology of `lime` 
read the [paper](https://arxiv.org/abs/1602.04938) and visit the repository of
the [original implementation](https://github.com/marcotcr/lime).

The `lime` package for R does not aim to be a line-by-line port of its Python
counterpart. Instead it takes the ideas laid out in the original code and 
implements them in an API that is idiomatic to R.

## An example
The only requirement for the classifier that `lime` imposes is that it must
implement a `predict()` method accepting a `type = 'prob'` argument (which 
outputs the probability for each class). Conveniently, this covers all 
classifiers available through [`caret`](https://CRAN.R-project.org/package=caret), 
and if your model of choice is not covered you have the possibility of writing
your own predict method.

The following shows how a random forest model is trained on the iris data set 
and how `lime` is then used to explain a set of new observations:

```{r, message=FALSE}
library(caret)
library(lime)

# Split up the data set
iris_test <- iris[1:5, 1:4]
iris_train <- iris[-(1:5), 1:4]
iris_lab <- iris[[5]][-(1:5)]

# Create Random Forest model on iris data
model <- train(iris_train, iris_lab, method = 'rf')

# Create explanation function
explain <- lime(iris_train, model)

# Explain new observation
explain(iris_test, n_labels = 1, n_features = 2)
```

## Installation
`lime` is still a work in progress and is thus not available on CRAN yet. In 
order to try it out install it directly from GitHub:

```{r, eval=FALSE}
# install.packages('devtools')
devtools::install_github('thomasp85/lime')
```

## Scope
In addition to standard tabular data the Python implementation also provides
specialized explainers for text and image data. Furthermore, the article also 
discusses how `lime` can be used for explaining *models* rather than 
*predictions*, using an approach called submodular picks. Lastly, there's 
obvious extensions to the approach such as using different algorithms for the 
local fit step, supporting regressors in addition to classifiers, and providing
additional feature selection algorithms. All of the above will hopefully find 
its way to this port in due time...
